# ABSTRACT

Nothing here but this must remain.

#### TODO
- [ ] update power analysis glossary
- [ ] add appendix figures
- [ ] edit or cut reference sites text


\newpage

```{r}
hbll_sites_all         <- readRDS(here::here("data-generated", "overlays", "hbll-site-summary-table.rds"))
hbll_n_site_rows       <- nrow(hbll_sites_all)
hbll_n_sites_unique    <- length(unique(hbll_sites_all$`Common Site Name`))

mean_fit_values <- readRDS(here::here("data-generated", "fit_characteristics.rds")) |>
  filter((species %in% c("canary rockfish", "silvergray rockfish") & survey %in% c("HBLL OUT N", "HBLL OUT S")) | !(species %in% c("canary rockfish", "silvergray rockfish"))) |>
  group_by(species) |>
  summarise(across(encountered_count_per_year:sigma_E, mean))

canary_encounter <- mean_fit_values |> 
  filter(species == "canary rockfish") |>
  pull(encountered_rate_restricted)

silvergray_encounter <- mean_fit_values |> 
  filter(species == "silvergray rockfish") |>
  pull(encountered_rate_restricted)

dogfish_encounter <- mean_fit_values |> 
  filter(species == "north pacific spiny dogfish") |>
  pull(encountered_rate_restricted)
```

# POWER ANALYSIS --- RECOMMENDATIONS FOR MONITORING PROGRAM DESIGN AND SAMPLING FREQUENCY

We selected the Hard Bottom Longline (HBLL) survey as a case study to evaluate sampling design requirements using power analysis and to assess how monitoring programs could be adapted or expanded to detect meaningful changes in conservation priority species.
Power analysis is a widely used approach for evaluating how sampling design and replication influence the ability to detect ecological change when it occurs.
Statistical power depends on the magnitude of the effect being tested, the degree of spatial and temporal variability in indicators, and the level of replication across sites, years, and sampling units [@underwoodBACISampling1994; @stewart-oatenEnvironmentalImpact1986].
In impact monitoring frameworks such as Before--After--Control--Impact (BACI) designs, balanced designs with comparable numbers of observations in pre- and post-impact periods generally provide greater statistical power than unbalanced designs with the same total sampling effort [@underwoodBACISampling1994; @smokorowskiCautionsUsing2017].

The HBLL survey was selected as an illustrative example for several reasons.
The HBLL scored high in our program assessment based on temporal suitability, domain relevance, and spatial coverage criteria. 
It has a 19-year time series providing sufficient historical data to characterize baseline variability, monitors multiple MPAN E-CP rockfish species (Yelloweye Rockfish, Quillback Rockfish), samples `r hbll_n_sites_unique` protected area sites with coverage both inside and outside MPAN boundaries, and collects detailed biological data (length, weight, archived otoliths) enabling analysis of multiple response metrics.

The power analyses conducted for the HBLL were used to inform recommendations on sampling frequency, spatial replication, and temporal coverage needed to support robust evaluation of MPA performance.
Importantly, this case study serves as one methodological example---not a prescription. 
The HBLL was not originally designed for MPA performance assessment; sampling locations were selected for regional stock assessment and conservation monitoring. 
The power analysis demonstrates what is currently feasible with existing data.
Different programs will require different adaptations depending on their design, species monitored, spatial scale, logistical constraints, and organizational priorities.
This case study illustrates a general analytical workflow that can be applied to other high-suitability programs, with final decisions about program adaptation to be made collaboratively by Network Partners through the Network Monitoring Strategy process.

## MONITORING PROGRAM DESIGN CONSIDERATIONS

**Statistical robustness in monitoring**

Effective sampling and monitoring design require statistical robustness, including careful consideration of sample size, replication, statistical power, and data independence. 
Neves et al. (in prep.) provides a comprehensive discussion of factors involved in monitoring program sampling design, including various sample design types, BACI approaches, and reference site selection (DFO 2021a).

**Replication**

Establishing appropriate replication levels, whether measured as the number of sites, years, or samples within sites, presents significant challenges and varies considerably among projects and indicators (Braun et al. 2019). 
Power analyses serve as valuable tools for understanding how replication influences statistical power, defined as the ability to detect effects when they exist. 
Statistical power depends on several key factors: spatial and temporal variation in indicators, the strength of relationships between indicators and responses (effect size), and replication or sample size, including the number of sites, years, and plots.

Power analyses can use standard methods for simple statistical analyses or simulations for more complex modelling approaches (Johnson et al. 2015). 
These analyses require assumptions about the spatial and temporal variability of indicators and about effect sizes specific to individual studies and indicators.

**Temporal considerations and study design**

Increasing the number of years of data collection generally enhances statistical power. 
Balanced designs with equal numbers of before and after years provide greater statistical power than unbalanced designs using the same total number of years (Braun et al. 2019). 
O\'Neal et al. (2016) recommend collecting a minimum of two years of before-project data, matched with two years of after-project data. 
While acquiring pre-project data can be challenging, the statistical benefits of a balanced design make a compelling case for investing in before-project data collection when feasible (Braun et al. 2019).

Within the NSB context, MPA sites experiencing greater pre-implementation fishing intensity and other pressures may have higher power to detect management-related changes than sites with historically lower human activity levels. 
This is because an ecological response is expected when fishing is reduced through protection (i.e., recovery), whereas in areas with low pre-implementation fishing pressure there may be little associated recovery because fishing pressure changes little before and after implementation.

**Before-After-Control-Impact (BACI) design**

A BACI design represents one of the most appropriate methods for evaluating management measure performance. 
This approach is particularly powerful because it controls for both temporal and spatial variation, thereby improving the robustness of conclusions regarding management effects. 
A BACI design compares sampling periods before and after measure implementation alongside impacted and control sites.

Successful BACI implementation requires suitable reference sites (control sites) that remain fixed throughout the study period, while within-site sampling units may be re-randomized or maintained as fixed locations. 
External reference sites are highly recommended to maximize the scope of inference and are required for BACI or similar sampling designs. 
These sites should be positioned sufficiently far from protected areas to avoid edge effects, maintain comparable environmental conditions and substrates, and exhibit similar historical pressure levels (DFO 2021a).

**Reference sites and monitoring**

Areas currently monitored outside protected areas serve as invaluable reference sites for assessing MPA performance and impact. 
These monitored zones, often sharing similar ecological characteristics with MPA areas, provide comparative benchmarks for evaluating changes and patterns within protected regions (Nature United 2023). 
Through examination of these external areas, scientists and conservationists can establish meaningful parallels, contrasting impacts of human activities and environmental changes in unprotected regions against conservation outcomes within MPAs.

Reference areas facilitate understanding of natural variations, human-induced pressures, and potential benefits derived from protective measures. 
Analysis of differences and similarities enables assessment of MPA efficacy, providing critical information for adaptive management and conservation strategy refinement within protected marine environments.

Reference monitoring represents an emerging opportunity that uses well-managed MPA sites as reference points for understanding the effects of global and regional drivers on ocean conditions (Fairweather 2012a; Fairweather 2012b). 
Because reference sites will develop over time, this approach has been identified as a low priority for the Network monitoring framework, strategy, and associated plans (Network Monitoring Framework). 
However, reference monitoring should be considered in subsequent iterations of Network monitoring after sufficient time has passed for site and network establishment.

```{r power-definitions}
readr::read_csv(here::here("text/data/power-definitions.csv")) |> 
  flextable() |>
  set_table_properties(layout = "autofit") |> 
  set_caption("Definitions used in power analysis.")
```

## POWER ANALYSIS METHODOLOGY

The HBLL survey consists of two survey series: inside (HBLL INS, referring to waters between Vancouver Island and the mainland) and outside (HBLL OUT, all other BC waters). 
North and south areas within each series are surveyed biennially. 
Three of the four regions (HBLL OUT N, HBLL OUT S, and HBLL INS N) overlap with the GBS MPA Network (Figure \@ref(fig:hbll-mpa-historical-overlay)).
The surveys use a depth-stratified random sampling design, with a target allocation of 189 sets (HBLL OUT N), 252 (HBLL OUT S), and 70 (HBLL INS N).

We used a simulation-based power analysis to assess the ability of the HBLL survey to detect changes in population indices of groundfish across the MPAN. 
Our approach consisted of four steps: (1) fit conditioning models to historical data to parameterise the operating model, (2) use the operating model to simulate population recovery trends inside the MPAs under the assumption of fishing cessation, (3) apply different sampling strategies to the simulated population, and (4) fit estimation (monitoring) models to the combined historical and simulated data to evaluate the power of these sampling strategies to detect trends across the MPAN.
This approach allows us to test different scenarios, such as varying recovery rates within MPAs or evaluation time frames, while incorporating some ecological complexity and variability.

Using a spatiotemporal modelling approach, we applied this power analysis to six conservation priority species that are well sampled by the HBLL surveys: Yelloweye Rockfish, Quillback Rockfish, Lingcod, Canary Rockfish, Silvergray Rockfish, and Pacific Spiny Dogfish. 
The HBLL surveys were designed to monitor Quillback Rockfish and Yelloweye Rockfish.
Lingcod and Pacific Spiny Dogfish are also well indexed by the HBLL surveys, and these surveys are used in their stock assessments.
Canary Rockfish and Silvergray Rockfish are generally found in deeper waters than Yelloweye or Quillback Rockfish and are encountered less frequently in the HBLL surveys than the other species considered here.
Their stock assessments use synoptic trawl survey data, not the HBLL survey data.
Furthermore, Canary and Silvergray Rockfish have not been encountered in the portion of HBLL INS that lies inside the MPAN.

### Fitting conditioning models

For each of the three HBLL survey regions and each species, we fit separate spatiotemporal models with the sdmTMB R package [@sdmTMB] of the form:

\begin{align}
  y_{\mathbf{s},t} &\sim \text{BetaBinomial}(n_{\mathbf{s},t}, \mu_{\mathbf{s},t}, \phi) \\
  \mu_{\mathbf{s},t} &= g^{-1}(\beta_t + \beta_\mathrm{MPA} + \omega_\mathbf{s} + \epsilon_{\mathbf{s},t}) \\
  \omega_\mathbf{s} &\sim \text{MVNormal}(\mathbf{0}, \mathbf{Q}_\omega^{-1}) \\
  \boldsymbol{\epsilon}_{t} &\sim \text{MVNormal}(\mathbf{0}, \mathbf{Q}_\epsilon^{-1})
\end{align}

We modelled the number of hooks catching a given species $y_{\mathbf{s},t}$ at each location $\mathbf{s}$ and year $t$.
We used a beta-binomial observation model, which is a generalization of the binomial distribution in which the probability of success is not fixed but varies according to a beta distribution.
This allows the model to capture additional variance across trials compared to the binomial distribution.
The symbol $n_{\mathbf{s},t}$ is the number of ``trials'' (here the number of hooks in a given survey set),
$\mu_{\mathbf{s},t}$ is the expected probability of a hook catching a fish, and
$\phi$ is the precision parameter, where smaller values of $\phi$ result in greater variance: $\mathrm{Var}(y) = n\mu(1-\mu)\frac{n+\phi}{1+\phi}$.
The function $g^{-1}$ represents the (inverse) complementary log-log ("cloglog") link chosen for its connection to the Poisson sampling process and its use in integrated data modelling.

The beta-binomial distribution was the most parsimonious observation model when compared using AIC against Poisson and negative binomial models fit to fish counts with log hook count as an offset.
Importantly, the beta-binomial constrained the number of fish observed to be within the number of hooks set and therefore produced simulated data that more closely resembled the observed data.
See supplementary materials for model comparison of hook and line count data simulation diagnostics.
We accounted for hook competition by scaling catch rates to reflect the number of hooks removed from fishing by fish during the set time. 
We used this adjusted hook count as the weight in our beta-binomial model [@anderson2019synopsis]: 

\[
\text{proportion of baited hooks} = \frac{\text{count bait only}}{\text{hook count}}
\]
\[
\text{hook adjustment factor} = \frac{-\ln(\text{proportion of baited hooks})}{1 - \text{proportion of baited hooks}}
\]
\[
\text{adjusted hook count} = \frac{\text{hook count}}{\text{hook adjustment factor}}
\]

Our linear predictor included annual independent year effects ($\beta_t$), an inside-vs.-outside MPA effect ($\beta_\mathrm{MPA}$), 
spatial random effects ($\omega_\mathbf{s}$), and spatiotemporal random effects ($\epsilon_{\mathbf{s},t}$).
These random effects were structured as Gaussian Markov random fields (GMRFs) with mean zero and precision matrices ($\mathbf{Q}_\omega$ and $\mathbf{Q}_\epsilon$; inverse covariance matrices) based on a Matérn spatial correlation function.
The spatial random fields account for persistent spatial patterning in distribution, while the spatiotemporal random fields account for spatial patterning that is unique to each year.
If models failed to converge or if spatial and/or spatiotemporal variance estimates were near zero ($<$ 0.01), the models were refit with the collapsed random fields set to zero.
Models were considered to have converged if the Hessian matrix was positive definite and the absolute gradient of the marginal log-likelihood with respect to each fixed effect was $<$ 0.001.

The spatial patterns predicted by the conditioning model varied by species (Figure \@ref(fig:predicted-distributions)).
Pacific Spiny Dogfish had the patchiest distribution, with strong hotspots masking patterns along the rest of the coast in the colour scale.
Parameter estimates from the conditioning models were relatively similar across surveys within a given species (Figure \@ref(fig:conditioning-coefs)).
The spatial range (distance at which two observations are effectively independent) ranged from about 50 km to about 100 km.
Interannual mean variation was positively correlated for Lingcod, Pacific Spiny Dogfish, and Quillback and was close to zero or negative in other cases.
Interannual mean variation and spatiotemporal variation were highest for Pacific Spiny Dogfish.
Spatial variation was strongest for Quillback Rockfish.
The inside vs. outside MPA parameter generally had a 95% confidence interval that crossed zero and would need to be interpreted alongside the spatial random field values, which are competing to explain the same variance.

### Simulating new observations

We then used the conditioning models to simulate future data with properties matching the historical data, but with a known underlying true recovery rate within the MPAs relative to the outside survey sites.
From the conditioning models we extracted the following parameters: the estimated spatial random field values, the spatial and spatiotemporal range, the beta-binomial precision parameter, the spatiotemporal random field standard deviation, the inside vs. outside MPA effect, and the annual mean estimates (year effects).
We fit a first-order autoregressive (AR1) model to the independent year effects estimated for each conditioning model (species-region combination) to estimate the temporal autocorrelation coefficient, $\rho_v$, and the marginal standard deviation, $\sigma_v$.

We then used these parameters to simulate future data with known recovery trends.
Our model matched the conditioning model in structure, with the following differences: 
First, future year mean values were randomly drawn from an AR1 time series with matching $\rho_v$ and $\sigma_v$ values.
This created annual mean values with similar variability and autocorrelation to the historical time period.
Second, a known recovery trend was imposed on sites within the MPAN.
We chose a sequence of possible recoveries over a 25-year time span, ranging from small recoveries to optimistic recoveries: 5%, 10%, 25%, and 50%.
These were parameterized as coefficients in the linear predictor, corresponding to coefficient values of, for example, log(1.10)/25 = 0.0038 for a 10% recovery over 25 years.
Sites outside the MPAN followed a flat trajectory over time on average.
Both the inside and outside sites were given additional noise around the mean trajectories from the AR1 time series noted above.

These true recovery trends were added to our simulation of the underlying spatial, spatiotemporal, and observation error properties.
In all cases, the historical data remained untouched and were used as they were observed when generating a complete historical and forward simulation dataset.
For a given simulation replicate of future data, we used the estimated spatial random field values (i.e., assuming a stationary distribution), drew a new set of spatiotemporal random fields for each year given the estimated variance and range, and drew a new set of observation error given the estimated beta-binomial precision.
We performed 160 replicates, which was sufficient for our power estimates to stabilize (Figure \@ref(fig:cumulative-power)).

We chose sampling locations for the future data by bootstrapping survey sites from the historical years.
After resampling years with replacement, we assigned all samples from a given historical year to a future year.
Our main experiment maintained this status quo sampling design, in which survey effort was unchanged.
We also explored, for one species as an example, an alternative reduced-sampling scheme in which inside-MPA sites were sampled only on every second survey (that is, every four years for these biennial surveys).
In that scenario, we resampled site locations from the existing outside-MPA sites in the corresponding year to redistribute the same survey effort.

We verified that our simulation models were able to generate observations that reflected the properties of our historical data through a sequence of visualizations.
First, to assess whether the interannual variation in the simulated data reflected the historical data, we visualized average observed catch rates per year inside and outside the MPAN in both time periods (Figure \@ref(fig:mean-observed-ts-eg)).
This also allows us to calibrate expectations for how challenging it may be to recover the underlying trends given the noise in the system.
We found that the two time periods looked similar in their time series properties.
Although the increase inside the MPAs is noticeable in some replicates, it is often obscured by temporal, spatiotemporal, and observation noise in the system.

Next, we evaluated whether the distribution of the observations was similar in the future versus historical time periods by plotting all observations separated by year.
We found that our simulation models were generally able to simulate data with a similar spread to the historical data (Figures A\@ref(fig:sim-posterior-yelloweye), A\@ref(fig:sim-posterior-lingcod), and A\@ref(fig:sim-posterior-quillback)).
This was not the case when we previously attempted other observation distributions such as the negative binomial, lognormal Poisson, or binomial.

Finally, we evaluated whether our simulation models could generate spatial observations with similar properties to the historical observed data.
Here, we asked whether one would be able to distinguish which panel was the "real" data and which were the simulation replicates if the panels were not labelled.
We generally found that it would be difficult to distinguish the real data from the simulated data when plotted spatially and therefore concluded that our simulation models could faithfully represent important properties of the historical data (Figures A\@ref(fig:turing-map-quillback), A\@ref(fig:turing-map-yelloweye), and A\@ref(fig:turing-map-lingcod)).

### Fitting estimation models to the simulated data

Finally, we fit estimation models to our simulated data with the goal of estimating the recovery rate inside the MPAs, using the outside sites and the historical time period as baseline references.
Our estimation model matched the conditioning model, with some exceptions to account for our inference goal.
We fit the historical years with the same independent (factor) year effects used in the conditioning model.
To estimate recovery, we added a predictor and associated coefficient that modelled a linear trend in log space over time for the inside-MPA sites across all future years.
We also included an intercept for this future time period.
Functionally, we did this by giving all future years a single "future year" factor level.
We explored an alternative estimation model with a first-order autoregressive, time-varying year effect across the full historical and future data span, thereby avoiding separate historical and future year factors.
We found that this autoregressive model gave similar results (not shown), and so we focus on the simpler model described above.
Our inference was based on whether the two-tailed 95% confidence interval for the parameter representing the inside-vs.-outside MPA by future year interaction crossed zero.

### Calculating power and magnitude error

From the interaction coefficient, we calculated several metrics related to statistical power.
First, we calculated correctly signed power.
Ordinarily, power is defined as the proportion of replicates with a significant effect in either direction when a true effect is present.
To avoid confusion, we define it here as the proportion of replicates in which abundance is assessed as significantly *increasing*.
The proportion of times that significant results have the wrong sign is often called a "sign" or "Type S" error [@gelman2014].
We found relatively strong sign errors in the early evaluation years, but do not focus on them here.
We also calculated magnitude, or exaggeration ratio, errors ("Type M" errors) [@gelman2014].
These represent the expected overestimate of the effect size when an effect is assessed as significant.
We calculated this by taking the average significant effect size and comparing it with the true effect size.
At low levels of power, these errors can be substantial [@gelman2014].
Specifically, we expressed these magnitude errors as a multiplicative factor on the estimated increase over 25 years.
For example, a magnitude error of two would mean that, if we detected a significant recovery estimated at 20%, the true underlying recovery would on average be only 10%.

## POWER ANALYSIS RESULTS

<!-- --- figure reference checklist to make sure all figures are referenced in the text
[x] hbll-mpa-historical-overlay.png (`hbll-mpa-historical-overlay`)
[ ] predicted-distributions.png (`predicted-distributions`)
[ ] mean-observed-ts-eg.png (`mean-observed-ts-eg`)
[x] main-power-plot.png (`power-plot`)
[x] type-m-error-plot.png (`type-m-plot`)
[x] power-correlations-25pct.png (`power-correlation-plot`)
[ ] yelloweye-power-bootstrap-vs-no-mpa-every-2nd-survey.png (`sampling-scenario-power-comparison`)

--- list of appendix figures:
[ ] conditioning-coefs.png (`conditioning-coefs`)
[ ] bias-check-on-estimate-lingcod.png (`bias-check-lingcod`)
[ ] cumulative-power-plot-all-species.png (`cumulative-power-all`)
[ ] sim-posterior-yelloweye rockfish-HBLL OUT N.png (`sim-posterior-ye-out-n`)
[ ] sim-posterior-lingcod-HBLL OUT N.png (`sim-posterior-lingcod-out-n`)
[ ] sim-posterior-quillback rockfish-HBLL OUT N.png (`sim-posterior-qb-out-n`)
[ ] turing-map-quillback rockfish-HBLL-OUT-N.png (`turing-map-qb-out-n`)
[ ] turing-map-yelloweye rockfish-HBLL-OUT-N.png (`turing-map-ye-out-n`)
[ ] turing-map-lingcod-HBLL-OUT-N.png (`turing-map-lingcod-out-n`)

[ ] power-phi-correlation.png (`power-phi-correlation`)
[ ] power-encounter-rate-correlation.png (`power-encounter-rate-correlation`)
[ ] power-sigmaE-correlation.png (`power-sigma-e-correlation`)
[ ] power-sigmaV-correlation.png (`power-sigma-v-correlation`)
-->

Almost 31\% of HBLL survey sets have occurred within the MPAN boundaries since 2003 (Figure \@ref(fig:hbll-mpa-historical-overlay)). 
Most sets in the MPAN are from the HBLL OUT N and HBLL OUT S regions.
Because the survey does not sample the existing RCA sites, relative MPAN coverage in HBLL INS N is much lower than in the other HBLL surveys.

```{r}
power_df <- readRDS(here::here('data-generated', 'power-df-historical-sampling.rds'))
power_wide <- power_df |>
  select(species, mpa_effect_label, eval_year, power_signed) |>
  tidyr::pivot_wider(names_from = "eval_year", values_from = "power_signed")

qb_25_power <- power_wide |>
  filter(species == "quillback rockfish", mpa_effect_label == "25%") |>
  mutate(across(where(is.numeric), \(x) sprintf("%.2f", round(x, 2))))
ling_25_power <- power_wide |>
  filter(species == "lingcod", mpa_effect_label == "25%") |>
  mutate(across(where(is.numeric), \(x) sprintf("%.2f", round(x, 2))))
ye_25_power <- power_wide |>
  filter(species == "yelloweye rockfish", mpa_effect_label == "25%") |>
  mutate(across(where(is.numeric), \(x) sprintf("%.2f", round(x, 2))))
ye_50_power <- power_wide |>
  filter(species == "yelloweye rockfish", mpa_effect_label == "50%") |>
  mutate(across(where(is.numeric), \(x) sprintf("%.2f", round(x, 2))))

type_m_wide <- power_wide <- power_df |>
  select(species, mpa_effect_label, eval_year, type_m_error) |>
  tidyr::pivot_wider(names_from = "eval_year", values_from = "type_m_error")

qb_25_type_m <- power_wide |> 
  filter(species == "quillback rockfish", mpa_effect_label == "25%")

rates_lu <- distinct(power_df, mpa_effect_label, true_effect) |>
  mutate(annual_rate = round(true_effect * 100, 1))

max_annual_rate <- max(rates_lu$annual_rate)
```

Our estimation models were generally able to recover true rates of recovery without bias (Figure B\@ref(fig:bias-check)). 
Estimates of the recovery trend became more accurate and less noisy across replicates after evaluating the MPAN for more years (Figure B\@ref(fig:bias-check)).
Our estimates of power had stabilized after running 160 replicates (Figure B\@ref(fig:cumulative-power)).

Power to detect recovery of abundance inside MPAs increased with longer evaluation timeframes (Figure \@ref(fig:power-plot)). 
For the first 13 years (2030, 2034, and 2038), regardless of the recovery effect size, power did not reach the target power of 80% for any species.
With 5% or 10% increases in abundance over 25 years, 80% power was never achieved for any species.
In our simulated system, we would not be able to statistically detect increases of only 5% or 10% for these species with these surveys, even after approximately 20 years.P
With 25% recoveries in abundance, target power was almost achieved after 20 years for three species: Quillback Rockfish (`r pull(qb_25_power, "2046")`), Lingcod (`r pull(ling_25_power, "2046")`), and Yelloweye Rockfish (`r pull(ye_25_power, "2046")`).
For 50\% recoveries in abundance, target power was achieved for all species after 20 years (Figure \@ref(fig:power-plot)).
For Quillback Rockfish, Lingcod, and Yelloweye Rockfish, target power for a 50% recovery in abundance was reached four years earlier, after 16 years (2042), than for the other three species. 

When power was low, magnitude errors or exaggeration rates were high (Figure \@ref(fig:type-m-plot)). 
For instance, if a significant recovery in abundance was detected in 2030, it was on average overestimated by a factor of 5 to 30 relative to the true rate.
By 2042, for the species with the highest power and for true recovery effect sizes of 25% or 50%, the exaggeration rates fell to less than twice the true rate, with the exception of Lingcod.
For example, given a true 25% recovery in Quillback Rockfish abundance over 25 years, if a significant positive effect was detected in 2038, the estimated recovery would on average be exaggerated by a factor of $\sim 2$ (i.e., estimated as a $\sim 50\%$ recovery).  

Power was positively correlated with encounter rates, but negatively correlated with interannual variability and spatiotemporal variability (Figure \@ref(fig:power-correlation-plot)).
Canary Rockfish and Silvergray Rockfish were the two species with the lowest encounter rates (`r round(canary_encounter * 100, 0)`\% and `r round(silvergray_encounter * 100, 0)`\% respectively).
Although Pacific Spiny Dogfish had the highest encounter rates (`r round(dogfish_encounter * 100, 0)`\%), its power was similarly low to that of Canary and Silvergray Rockfish, potentially due to high interannual variability, high spatiotemporal variability (Figure \@ref(fig:power-correlation-plot)), and high observation error (Figure B\@ref(fig:power-phi-correlation)).

When MPA sampling frequency was reduced to every fourth year, compared to the biennial status quo sampling, power was lower for the three species considered (Yelloweye Rockfish, Lingcod, and Quillback Rockfish; Figure \@ref(fig:sampling-scenario-power-comparison)).
Reduced sampling delayed when a given power level could be reached by approximately four years. 
This delay also occurred in the exaggeration ratios, with higher exaggeration ratios when MPA sampling frequency was reduced. 

# Discussion

Statistical power to detect MPA recovery trends varied across species and effect sizes.
For species with faster recovery rates, ≥80% power could be reached within the first 10--15 years of MPA implementation using status quo sampling, while for species with slower recovery it may take 20--25 years of monitoring.
Reducing MPA sampling frequency generally delayed when ≥80% power was reached by approximately four years (sampling cycle).


Extensions:

- AR1 on year effects, but made little difference

- increased sampling scenario

- could do by survey - would very likely be weaker still

- could combine with other surveys such as the synoptic trawl - more power!


- what are reasonable rates of recovery by stock ... lingcod at B0... yelloweye, quillback - well below B0 but late age at maturity long generation time... expect slower recovery

- and we did a exponential growth recovery trend - would actually level off towards B0 *and* rockfish often have spasmodic recruitment... so for numbers, very little in most years, occasionally large recruitment events - we did a continuous linear effect - probably easier to detect what we did!

* We assumed a constant linear rate of increase over the 25 years. Although the recovery rates calculated from the stock assessments reflect current levels of depletion and density dependent effects. 

- dogfish - slow recovery + movement may disperse them

- will depend on movement too

* This analysis assumed no species movement.
Any potential spillover effects or capturing of fishes that move outside of the MPA boundaries would reduce recovery rates and our ability to detect MPA effects.

- body size may be faster - reference Kyle's preprint
  - growth
  - biomass will be faster
  - HBLL is abundance - reference the Perkins et al. paper

* Recent analyses have shown that body size rather than abundance is more sensitive to detection. [@wilson2025groundfish; @perkins2021quantifying]


Helps us set expectations:

- At best expect detection of MPA effects in the best sampled species within 10-15 years

- Rates of recovery \~ 1% per year expect to detect effects in 20-25 years

- Decreasing sampling within MPAs to every 4 years increased the number of years before we could detect an effect by \~ 5-10 years for a given recovery rate

Detection of significant effects before 15 years is likely to lead to over estimates of effects 🡪 high exaggeration rates.

Stock assessment "no fishing" scenarios and life histories can help to inform expectations of recovery rates

**Pre-implementation fishing pressure**
Pre-implementation fishing pressure will affect the recovery rates -- i.e., sites that had low or no pre-implementation fishing pressure are not expected to recover. 
Similarly, only sites that have have interventions that reduce or exclude fishing pressure are expected to recover.
Therefore, if MPA effects are assessed across all sites in the network, power will lower than simulated here.


**Morris et al. 2024 in relation to our work**
- Tested much larger effect sizes than we did - specifically they were considering these changes over much shorter time frames as well: 3- 10 years

- Status quo monitoring (23 trawls / stratum) could detect when a 20% decline per year for smooth skate. Compared to our maximum rate with a 50% increase in abundance over 25 years, which is a `r max_annual_rate` \% increase per year.

- For black dogfish, the most frequently encountered and highest CPUE species they considered 

- They did find that the lowest CPUE species (Northern Wolfish) had the lowest power 


---

**SAR / main text discussion bullet points**

- 

- 

- 


**Uncertainties/limitations (bullet points for main text discussion):** 

- 

- 